The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2011-11-28 smurf_paper.tex
Move step fixing details to an appendix

Sudden steps can occur in the time stream data from each bolometer.
Fig.~\ref{fig:steps1} shows such steps in the time streams for two
bolometers. For each bolometer, correcting these steps involves:

@@

\item Determining the widths of the candidate steps. Some steps
occur over a single sample such as the step close to sample 5000 in the
right-hand plot of Fig.~\ref{fig:steps1}, but others happen more
gradually such as the step close to sample 5300. In addition, a step can
be preceeded or followed by a short period of instability, as is visible
at the bottom of the step in the right hand plot of
Fig.~\ref{fig:steps1}.

@@

\paragraph{Detecting Candidate Steps:}
The first stage in the algorithm is to search the bolometer time stream for
contiguous blocks of samples that have unusually large mean gradients
compared to the neighbouring samples. Each such block of samples is
considered to be a candidate step. The starting and ending time of each
such block is noted. Once all candidate steps have been found for a
bolometer, the algorithm proceeds to the next stage. Candidate steps are
found as follows:

\item At each time sample, find the difference between the two adjacent
values in the median filtered bolometer data. This is proportional to the
gradient of the median filtered bolometer data. Fig.~\ref{fig:steps3}
shows these differences for the sample bolometer shown in previous
figures. The high gradients at the three step are clearly visible at
indices 5040, 5100 and 5280. In addition, the varying gradient of the
underlying noise-free time series can be seen in the lower level changes
in the difference values.

\item Steps are located by looking for difference values that exceed 25
times the RMS of the neighbouring difference value. However, the RMS of
the differences can be heavily affected by the gradient of the underlying
noise-free time series. If there is a strong upward or downward trend in
the time series data, the RMS of the difference values will be higher
potentially resulting in some steps being missed. Therefore to enhance
the reliability of the step detection, an attempt is made to remove these
variations before locating the steps. To do this, the differences are
smoothed using mean filter, and the smoothed differences are removed. The
remaining residual difference values represent the gradient due to steps,
spikes, noise, etc. It is important that the smoothed differences are not
polluted by the large values caused by spikes and steps. So before
smoothing, differences of more than three times the RMS difference (taken
over the whole time stream) are identified and flagged. The RMS is then
re-calculated omitting the flagged values and a further two such
rejection iterations are performed. This flags high gradient values
caused by spikes, jumps, point sources, etc. The difference values close
to a step are often not typical of those in the wider neighbourhood, so
each block of contiguous flagged difference values is doubled in width.
These flagged values are shown as black in Fig.~\ref{fig:steps4}, with
the remaining unflagged values shown in red.

The data in Fig.~\ref{fig:steps5} results in two blocks being defined -
one between samples 5037 and 5102, and another between samples 5286 and
5296. Each of these blocks define the start and end of a candidate step.

@@

%------------------------------------------------------------------------------
\appendix
\section[]{This is an appendix}
%------------------------------------------------------------------------------

>>>>

Sudden steps can occur in the time stream data from each bolometer.

\emph{why?}

The black curves in Fig.~\ref{fig:steps1} show examples of such steps in
the time streams for two bolometers. If not removed, these steps can
cause severe ringing in the FLT model (Section~\ref{sec:flt}) and visible
streaks in the final map, corresponding to the paths of individual
bolometers over the sky.

Steps occur with a wide range of heights and shapes. The ratio of step
height to noise can vary from less than 10 to several hundred. Some steps
occur over a single sample such as the step close to sample 5000 in the
right-hand plot of Fig.~\ref{fig:steps1}, but others happen more
gradually such as the step close to sample 5300. In addition, a step can
be preceeded or followed by a short period of instability, as is visible
at the bottom of the step in the left hand plot of
Fig.~\ref{fig:steps1}. Further problems are caused by steps that occur
close together in time, such as the large downward step followed by a
smaller upward step close to sample 5000 in the right hand plot of
Fig.~\ref{fig:steps1}.

Detecting and correcting such a wide variety of steps reliably has proved
to be a challenge. The current step-correction algorithm is described in
detail in appendix~\ref{appendix:steps}. In outline, the following stages
are involved in detecting and correcting steps in a single bolometer time
stream:

@@

\item Median smooth the whole time stream.
\item Find the gradient of the median smoothed time series at each sample.
\item Smooth the gradient values to determine the local mean gradient and
subtract this local mean from the total gradient to get residual gradients.
\item Find residual gradient values that exceed 25 times the local RMS of
the residual gradients.
\item Group these high residual gradients into contiguous blocks of
samples.
\item Merge blocks that are separated by less than 100 samples

@@

The results of step correction are shown by the red curves in
Fig.~\ref{fig:steps1}. For comparison, the blue curve shows the
uncorrected time stream from a nearby bolometer that does not suffer from
steps. The agreement between the red and blue curves confirms that
the step correction algorithm is working satisfactorily.

@@

%------------------------------------------------------------------------------
\appendix
\section[]{Step correction in detail}

An outline of the step correction algorithm is given in
Section~\ref{sec:steps}. This appendix contains a more detailed
description.

Detecting and correcting steps in a single bolometer time stream involves:

\item Determining the widths of the candidate steps.

\subsection{Detecting candidate steps}
The first stage in the algorithm is to search the bolometer time stream for
contiguous blocks of samples that have unusually large mean gradients
compared to the neighbouring samples. Each such block of samples is
considered to be a candidate step. The starting and ending time of each
such block is noted. Once all candidate steps have been found for a
bolometer, the algorithm proceeds to the next stage. Candidate steps are
found as follows:

\item At each time sample, find the difference between the two adjacent
values in the median filtered bolometer data. This is proportional to the
gradient of the median filtered bolometer data. Fig.~\ref{fig:steps3}
shows these differences for the sample bolometer shown in previous
figures. The high gradients at the three step are clearly visible at
indices 5040, 5100 and 5280.

\item Steps are located by looking for difference values that exceed 25
times the RMS value of the nearby neighbouring differences. However, the
local RMS of the differences includes the gradient of the underlying
noise-free time series. So if there is a strong upward or downward trend
in the time series data, the RMS of the difference values will be higher,
potentially resulting in some steps being missed. For instance, at sample
5000 the overall trend in the time series is downwards, resulting in a
negative bias to the difference values.\footnote{This bias is not visible in
Fig.~\ref{fig:steps3} because of the large vertical range, but can be seen in
Fig.~\ref{fig:steps4} which has a smaller vertical scale.}

Therefore to enhance the reliability of the step detection, such biases
are removed before locating the steps. To do this, the differences are
smoothed using a mean filter, and the smoothed differences are removed.
The remaining residual difference values represent the gradient due to
steps, spikes, noise, etc. It is important that the smoothed differences
are not polluted by the large values caused by spikes and steps. So
before smoothing, differences of more than three times the RMS difference
(taken over the whole time stream) are identified and flagged. The RMS is
then re-calculated omitting the flagged values and a further two such
rejection iterations are performed. This process flags high gradient values
caused by spikes, jumps, point sources, etc. The difference values close
to a step are often not typical of those in the wider neighbourhood, so
each block of contiguous flagged difference values is doubled in width.
These flagged values are shown as black in Fig.~\ref{fig:steps4}, with
the remaining unflagged values shown in red.

The data in Fig.~\ref{fig:steps5} results in two blocks being defined -
one between samples 5037 and 5102 and another between samples 5286 and
5296. Each of these blocks define the start and end of a candidate step.
The first block includes two visible steps, but they have been merged
into a single candidate step due to their small separation (less than 100
samples).

\subsection{Measuring and confirming candidate steps}

%------------------------------------------------------------------------------
